---
title: "GO enrichment analysis using `rGREAT` on non-coding accelerated elements"
author: Paula Beati
date: "April 7, 2025"
output:
  html_document:
    number_sections: false
    css: "style.css"
---

# Go term analysis

In this document you will follow the pipeline used to compute GO term enrichment for non-coding accelerated regions.

## Software

rGREAT <https://github.com/jokergoo/rGREAT>

## Input data

### Mammals (ncMARs)

Coordinates: hg38

Conserved elements: `./data/input/10_mammals_conserved_noncoding_elements.bed`

Accelerated elements: `./data/input/13_non_coding_acc_regions_mammals_ranked.bed`

### Aves (ncAvARs)

Coordinates: galGal6

Conserved elements: `./data/input/12_aves_conserved_noncoding_elements.bed`

Accelerated elements: `./data/input/14_non_coding_acc_regions_aves_ranked.bed`

### Common accelerated regions (CARs)

Coordinates: hg38

Conserved elements: `./data/input/10_mammals_conserved_noncoding_elements.bed`

Accelerated elements: `./data/input/supp14_ncCAR_hg38.bed`

## Pipeline

### Includes

```{r includes}

base_path <<- getwd()
script_base_path <<- file.path(base_path, 'source')
data_base_path <<- file.path(base_path, 'data')
go_terms_data_base_path <<- file.path(data_base_path, 'output', 'go_terms')

source(file.path(script_base_path,  'rGreat_base.R'))
source(file.path(script_base_path, 'rGreat_simulation.R'))
source(file.path(script_base_path, 'rGreat_analysis.R'))
source(file.path(script_base_path, 'rGREAT_regions_in_genes.R'))
source(file.path(script_base_path, 'rGreat_results.R'))
source(file.path(script_base_path, 'dotplot.R'))

```

Steps 1 to 3 are performed in function `main_mammals` in `rGreat_analysis.R`

### Step 1

Load or generate `rGREAT` evaluation for random sets of non-coding conserved elements, <GO:BP> ontology.

```{r simulations, echo = T, results = 'hide'}
    go_ontology <- 'bp'

    nc_cons_mammals_file_name <- '10_mammals_conserved_noncoding_elements.bed'
    nc_cons_mammals_file_path <- file.path(data_base_path, 'input', nc_cons_mammals_file_name)
    file.exists(nc_cons_mammals_file_path)
    nc_cons_mammals <- read.delim(nc_cons_mammals_file_path, sep = ' ', header = TRUE)

    nc_acc_mammals_file_name <- '13_non_coding_acc_regions_mammals_ranked.bed'
    nc_acc_mammals_file_path <- file.path(data_base_path, 'input', nc_acc_mammals_file_name)
    file.exists(nc_acc_mammals_file_path)
    nc_acc_mammals <- read.delim(nc_acc_mammals_file_path, sep = ' ', header = TRUE)

    hits_stats_sim <- simulation_stats_clade_go_terms(nc_cons_mammals,
                                        nc_acc_mammals, go_ontology, 'mammals')
```

### Step 2

Evaluate `rGREAT` on non-coding accelerated elements.

```{r rGREAT_acc, message = FALSE}
    # 2. calculate rgreat for acc elements
    go_terms_nc_acc_mammals_all <- enrichment_clade_go_terms(nc_acc_mammals,
                                            go_ontology, 'mammals', 'nc_acc_')

```

### Step 3

Filter significant GO terms from `rGREAT` results on non-coding accelerated elements.

```{r significance, message = FALSE}
    # 3. empirical p values for acc elements
    significant_clade_go_terms(go_terms_nc_acc_mammals_all, hits_stats_sim,
                                   go_ontology, 'mammals', 'nc_acc_')

```

### Step 4

Associate regions to genes using `rGREAT`

The following code is part of the function `main_mammals` in `rGREAT_regions_in_genes.R`

```{r gene_annotation, echo = T, results = 'hide', message = FALSE}

    clade <- 'mammals'
    ontology <- 'bp'

    go_annotation_type <- 'biomart'
    result_file_path <- bind_genes_data(clade, ontology, go_annotation_type)
    result_data_biomart <- read.delim(result_file_path, sep = '\t', header = TRUE)

    go_annotation_type <- 'default'
    result_file_path <- bind_genes_data(clade, ontology, go_annotation_type)
    result_data_default <- read.delim(result_file_path, sep = '\t', header = TRUE)


```

### Step 5

Format result tables.

The following code is part of the function `main` in `rGreat_results.R`

```{r format}
    clade <- 'mammals'
    ontology <- 'bp'
    
    data_file_path_1 <- combine_results_regions(clade, ontology)
    data_1 <- read.delim(data_file_path_1, sep = '\t', header = TRUE)

    data_file_path_2 <- combine_results_annotations(clade, ontology)
    data_2 <- read.delim(data_file_path_2, sep = '\t', header = TRUE)

    print(nrow(data_1))
    print(nrow(data_2))
    print(nrow(data_1) == nrow(data_2))

```

### Step 6

Represent top 30 GO enrichemnt result in a dot plot.

The following code is part of the function `main` in `dotplot.R`

```{r dotplot}
    
    clade <- 'mammals'
    go_ontology <- 'bp'

    plot_to_save <- dot_plot_clade_ontology(go_terms_data_base_path, go_ontology, clade)
    save_plot(plot_to_save, go_ontology, clade)

```

### Result

```{r show, out.width="100%", out.height="500px"}

    plot_base_path <- file.path(go_terms_data_base_path, clade, 'dot_plot')
    plot_file_name <- paste0('acc_nc_', clade, '_go', go_ontology, 
                             '_top30fe.pdf')
    plot_file_path <- file.path(plot_base_path, plot_file_name)
    
    knitr::include_graphics(plot_file_path) 
```
