---
title: "AnalysisElements"
author: "Anabella Trigila"
date: "7/6/2020"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(IRanges)
library(GenomicRanges)
library(ggplot2)

base_path <- '/u01/home/pbeati/2024/lucia/paper_acelerados/ncomms-24-47757/downstream'
data_base_path <- file.path(base_path, 'data')
out_base_path <- file.path(base_path, 'outputs')
scripts_base_path <- file.path(base_path, 'code')

source(file.path(scripts_base_path, "plot_scatter.R"))
source(file.path(scripts_base_path, "plot_bargraph.R"))

```

# Load objects

```{r load objects}


annotations_base_path <- file.path(data_base_path, 
                        "5. mammals accelerated candidate functional regions")
annotations_file_path <- file.path(annotations_base_path, 
                                   "CodingExonsRefSeqAll.hg38.bed")

if(! file.exists(annotations_file_path)) {
    annotations_file_path_zip <- file.path(annotations_base_path, 
                        "CodingExonsRefSeqAll.hg38.zip")
    unzip(zipfile = annotations_file_path_zip, 
          exdir = annotations_base_path)
}

refseq.exons.mammals <- read.delim(annotations_file_path, header = FALSE) %>% 
    dplyr::select(V1,V2,V3)

if(file.exists(annotations_file_path)) {
    file.remove(annotations_file_path)    
}

refseq_exons_mammals_granges <- makeGRangesFromDataFrame(refseq.exons.mammals,
                                                         seqnames.field = "V1",
                                                         start.field = "V2",
                                                         end.field = "V3")
genome(refseq_exons_mammals_granges) <- "hg38"

# -----

annotations_base_path <- file.path(data_base_path, 
                        "6. aves accelerated candidate functional regions")
annotations_file_path <- file.path(annotations_base_path, 
                                   "CodingExonsRefSeqAll.galGal6.bed")

if(! file.exists(annotations_file_path)) {
    annotations_file_path_zip <- file.path(annotations_base_path, 
                        "CodingExonsRefSeqAll.galGal6.zip")
    unzip(zipfile = annotations_file_path_zip, 
          exdir = annotations_base_path)
}

refseq.exons.chicken <- read.delim(annotations_file_path, header = FALSE)  %>% 
    dplyr::select(V1,V2,V3) 

if(file.exists(annotations_file_path)) {
    file.remove(annotations_file_path)    
}

# Aves exons
refseq_exons_aves_granges <- makeGRangesFromDataFrame(refseq.exons.chicken,
                                                      seqnames.field = "V1",
                                                      start.field = "V2",
                                                      end.field = "V3")
genome(refseq_exons_aves_granges) <- "galGal6"


# -----

mammals_acc_path <- file.path(out_base_path, 
                                  '3_mammals_candidate_func_elements.bed')
mammals_acc_ranges_df <- read.table(mammals_acc_path)
mammals_acc_ranges <- makeGRangesFromDataFrame(mammals_acc_ranges_df, 
                                               keep.extra.columns = TRUE)
mammals_acc_coding <- subsetByOverlaps(mammals_acc_ranges,
                                       refseq_exons_mammals_granges) 

#---

aves_acc_path <- file.path(out_base_path, 
                               "4_aves_candidate_func_elements.bed")     

aves_acc_ranges_df <- read.table(aves_acc_path)
aves_acc_ranges <- makeGRangesFromDataFrame(aves_acc_ranges_df, 
                                               keep.extra.columns = TRUE)

aves_acc_coding <- subsetByOverlaps(aves_acc_ranges,
                                    refseq_exons_aves_granges) 


#------
annotations_base_path <- file.path(data_base_path, 
                                   "7. human chicken transcripts")

human_genes_file_path <- file.path(annotations_base_path, 
                                   "knownCanonical(hg38).bed")
if(!file.exists(human_genes_file_path)) {
    annotations_file_path_zip <- file.path(annotations_base_path, 
                                   "knownCanonical(hg38).zip")
    unzip(zipfile = annotations_file_path_zip, 
          exdir = annotations_base_path)
}

human_genes <- read.table(human_genes_file_path) %>%
  dplyr::rename("chr" = "V1", "start" = "V2", "end" = "V3", "name" = "V4") %>%
  mutate(length = end-start)

#clean unzipped file
if(file.exists(human_genes_file_path)) {
    file.remove(human_genes_file_path)
}

human_genes_granges <- human_genes %>%
  makeGRangesFromDataFrame(keep.extra.columns=TRUE)


#------

annotations_base_path <- file.path(data_base_path, 
                                   "7. human chicken transcripts")

avian_genes_file_path <- file.path(annotations_base_path, 
                                   "galGal6_transcripts.bed")
if(!file.exists(avian_genes_file_path)) {
    annotations_file_path_zip <- file.path(annotations_base_path, 
                                   "galGal6_transcripts.zip")
    unzip(zipfile = annotations_file_path_zip, 
          exdir = annotations_base_path)

}

# Load chicken gene regions
avian_genes <- read.table(avian_genes_file_path) 
colnames(avian_genes) <- c("seqname", "start", "end", "name")
#clean unzipped file
if(file.exists(avian_genes_file_path)) {
    file.remove(avian_genes_file_path)
}
avian_genes_granges <- makeGRangesFromDataFrame(avian_genes, 
                                                keep.extra.columns=TRUE)
genome(avian_genes_granges) <- "galGal6"


```



# Coding accelerated elements (cMARs & cAvARs)

We assigned a name to each coding accelerated element based on a descending ranking of the number of changes ('shifts').

```{r raking coding MARs and AvARs}

# Assign names to Mammal coding accelerated elements (cMARs) based on their coding acceleration
acc_regions_mammals_file_path <- file.path(data_base_path, 
                                "5. mammals accelerated candidate functional regions", 
                                "mammals_acc_candidate_func_elements_all.txt")
acc_regions_mammals <- read.delim(acc_regions_mammals_file_path)
                                                         
coding_acc_regions_mammals_ranked <- acc_regions_mammals %>%
    select(phastCons_seqnames, phastCons_start, 
         phastCons_end, phastCons_id,
         functional_element_shift_count_exclgt0gap) %>%
  unique.data.frame() %>%
  inner_join(as.data.frame(mammals_acc_coding),
             by = c("phastCons_seqnames"="seqnames",
                    "phastCons_start" = "start",
                    "phastCons_end" = "end")) %>%
  arrange(desc(functional_element_shift_count_exclgt0gap)) %>%
  mutate(element_accelerated_ranking = row_number(),
         element_accelerated_ranking_name = paste0("cMAR-", row_number())) %>%
  dplyr::select(phastCons_seqnames,
                phastCons_start,
                phastCons_end,
                element_accelerated_ranking_name) %>%
  makeGRangesFromDataFrame(keep.extra.columns=TRUE)

# Assign names to Avian coding accelerated elements (cAvARs) based on their coding acceleration

acc_regions_aves_file_path <- file.path(data_base_path, 
                                "6. aves accelerated candidate functional regions", 
                                "aves_acc_candidate_func_elements_all.txt")
acc_regions_aves <- read.delim(acc_regions_aves_file_path)


coding_acc_regions_aves_ranked <- acc_regions_aves %>%
  select(phastCons_seqnames, phastCons_start,
         phastCons_end, phastCons_id,
         functional_element_shift_count_exclgt0gap) %>%
  unique.data.frame() %>%
  inner_join(as.data.frame(aves_acc_coding),
             by = c("phastCons_seqnames"="seqnames",
                    "phastCons_start" = "start",
                    "phastCons_end" = "end")) %>% arrange(desc(functional_element_shift_count_exclgt0gap)) %>%
  mutate(element_accelerated_ranking = row_number(),
         element_accelerated_ranking_name = paste0("cAvAR-", row_number())) %>%
  dplyr::select(phastCons_seqnames,
                phastCons_start,
                phastCons_end,
                element_accelerated_ranking_name) %>%
  makeGRangesFromDataFrame(keep.extra.columns=TRUE)

# Write to outputs

# Mammal cMARs
coding_acc_regions_mammals_ranked_out_path <- file.path(out_base_path, 
                                    "31_coding_acc_regions_mammals_ranked.bed")
write.table(coding_acc_regions_mammals_ranked, 
            file = coding_acc_regions_mammals_ranked_out_path)
# Avian cAvARs
coding_acc_regions_aves_ranked_out_path <- file.path(out_base_path, 
                                    "32_coding_acc_regions_aves_ranked.bed")
write.table(coding_acc_regions_aves_ranked, 
            file = coding_acc_regions_aves_ranked_out_path)


```

## Ranking of genes with the most coding accelerated regions

We ranked genes by their number of coding accelerated regions by intersecting the previously calculated coding accelerated elements and the hg38 and galgal6 coding genes ranges.

```{r genes with coding MARs and AvARs}

# Intersection of Coding Mammal accelerated functional elements with Human Genes
mammal_genes_with_cMARs_overlaps <- findOverlaps(human_genes_granges,
                                                 coding_acc_regions_mammals_ranked) 

# Sorting genes by number of cMARs
mammal_genes_ranked_coding <- as.data.frame(human_genes_granges[queryHits(mammal_genes_with_cMARs_overlaps),]) %>%
  group_by(name) %>%
  summarize(number_of_MARs =  n()) %>%
  arrange(desc(number_of_MARs))

# Intersection of Coding Avian accelerated functional elements with Chicken Genes
aves_genes_with_cAvARs_overlaps <- findOverlaps(avian_genes_granges,
                                                coding_acc_regions_aves_ranked) 

# Sorting genes by number of cAvARs
avian_genes_ranked_coding <- as.data.frame(avian_genes_granges[queryHits(aves_genes_with_cAvARs_overlaps),]) %>%
  group_by(name) %>%
  summarize(number_of_cAvARs =  n()) %>%
  arrange(desc(number_of_cAvARs))


# Plot


barplot_mammal_coding <- plot_horizontal_bar(
    head(mammal_genes_ranked_coding, n = 20), 
    number_of_MARs,
    name, "cMARs",
    file.path(scripts_base_path, "mammal.png"))

barplot_aves_coding <- plot_horizontal_bar(
    head(avian_genes_ranked_coding, n = 20), 
    number_of_cAvARs,
    name, "cAvARs",
    file.path(scripts_base_path, "bird.png"))


```

We then produced plots of gene with the highest number of coding accelerated elements and also the distribution based on the number of exons per gene.

```{r exons per gen mammals, fig.height=11.7, fig.width=8.3}

# Find overlaps between human genes and refseq exons -> This shows the number of exons per gene
exons_per_gene <- findOverlaps(human_genes_granges,
                               refseq_exons_mammals_granges)

# Count the number of exons per gene
exons_gene_coding <- as.data.frame(human_genes_granges[queryHits(exons_per_gene),]) %>%
  group_by(name) %>%
  summarize(number_of_exons =  n()) %>%
  arrange(desc(number_of_exons)) # Number of exons includes alternative exons

# Create a table that has the number of exons per gene and the numer of cMARs
table_mammals_perTU_coding <- inner_join(mammal_genes_ranked_coding,
                                         exons_gene_coding, by = 'name')

# Plot

scatter_plot <- plot_scatter(
  data = table_mammals_perTU_coding,
  x_var = "number_of_exons",
  y_var = "number_of_MARs",
  name_var = "name",
  x_label = "No. Exons",
  y_label = "No. cMARs",
  png_image =  file.path(scripts_base_path, "mammal.png")
) 


p1 <- cowplot::plot_grid(barplot_mammal_coding, barplot_aves_coding, 
                         labels = c('A', 'B'),
                         label_size = 12, ncol = 2,
                         align = "h", axis = "b")


p2 <- cowplot::plot_grid(p1, scatter_plot,
                         labels = c('A', ''),
                         label_size = 12,
                         ncol = 1)

# Export plot to outputs
ggplot2::ggsave(p2,
        filename = file.path(out_base_path, "33_coding_regions_genes.pdf"),
        width = 5, height = 10, units = "in")

```

Finally, we calculated if these cMARs and cAvARs are clustered and how is their distribution.

```{r coding clusters cMARs and cAvARs}

# Calculate clustered cMARs 
coding_acc_regions_mammals_ranked$cluster <- subjectHits(
    findOverlaps(coding_acc_regions_mammals_ranked, 
                 reduce(coding_acc_regions_mammals_ranked, min.gapwidth = 50000L))) # Adds a new column to all accelerated elements classyfing each of them to a cluster

# Dataframe with cMARs in clusters
df_mammal_cMARs_clusters <- data.frame(
  name = coding_acc_regions_mammals_ranked$element_accelerated_ranking_name,
                        cluster = coding_acc_regions_mammals_ranked$cluster,
                        chr = as.character(coding_acc_regions_mammals_ranked@seqnames),
  start = as.character(coding_acc_regions_mammals_ranked@ranges@start),
  end = as.character(coding_acc_regions_mammals_ranked@ranges@start + coding_acc_regions_mammals_ranked@ranges@width)) %>%
  group_by(cluster) %>%
  mutate(freq = n()) %>%
  arrange(desc(freq), cluster) %>%
  filter(freq > 3)

# Calculate clustered cAvARs 
coding_acc_regions_aves_ranked$cluster <- subjectHits(findOverlaps(coding_acc_regions_aves_ranked, reduce(coding_acc_regions_aves_ranked, min.gapwidth = 50000L))) # Adds a new column to all accelerated elements classyfing each of them to a cluster

# Dataframe with cAvARs in clusters
df_aves_cAvARs_clusters <- data.frame(
  name = coding_acc_regions_aves_ranked$element_accelerated_ranking_name,
                        cluster = coding_acc_regions_aves_ranked$cluster,
                        chr = as.character(coding_acc_regions_aves_ranked@seqnames),
  start = as.character(coding_acc_regions_aves_ranked@ranges@start),
  end = as.character(coding_acc_regions_aves_ranked@ranges@start + coding_acc_regions_aves_ranked@ranges@width)) %>%
  group_by(cluster) %>%
  mutate(freq = n()) %>%
  arrange(desc(freq), cluster) %>%
  filter(freq > 3)


# Export table to outputs
write.table(df_mammal_cMARs_clusters,
            file = file.path(out_base_path, "34_coding_cMARs_clusters.txt"),
            row.names = FALSE, quote = FALSE)
# Avian cAvARs
write.table(df_aves_cAvARs_clusters,
            file = file.path(out_base_path, "35_coding_cAvARs_clusters.txt"),
            row.names = FALSE, quote = FALSE)
```

# Run phastBias

Detect accelerated regions undergoing GC-biased gene conversion to differentiate from positive selection events.This is run in a cluster because it uses MAF files.

```{r eval=FALSE}
# Run phastBias (should be done in a cluster/cloud)
phastbias_file_path <- file.path(scripts_base_path, "run_phastbias.R")
source(phastbias_file_path)
```
