---
title: "Example of acceleration in mammals"
author: "Paula Beati"
date: "2025-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Acceleration in mammals 

The aim of this example is to calculate acceleration in a sample multiple alignment file (chr114, chr14sample.maf.gz), derived from the MAF file for chromosome 14 available at <https://hgdownload.soe.ucsc.edu/goldenPath/hg38/multiz100way/maf/chr14.maf.gz>,

## Setup

`cladeAcc` is a package developed to implement the pipeline steps for calculating accelerated elements.
It is available at <https://github.com/paulati/cladeAcc>

We need to install it

```{r pack_setup}

if (!require(rphast, quietly = TRUE))
    devtools::install_github("CshlSiepelLab/RPHAST")

if (!require("cladeAcc", quietly = TRUE))
    devtools::install_github(repo = 'https://github.com/paulati/cladeAcc')

```


## Local setup

### User defined paths 

A custom path `user_base_path` must be defined to store temporary results and pipeline output. If not specified, the data will be saved to the default location: ~/.local/share/R/cladeAcc/.

Note 1: Use <<- to assign this value globally (required).
Note 2: Use the full path; do not use ~ as a shortcut for the home directory (e.g., use /home/user/... instead). The ~ symbol may not be resolved correctly.

`git_base_path` refers to the location where this repository was cloned.

```{r user_defined_paths}

user_base_path <<- '/u01/home/pbeati/2024/lucia/paper_acelerados/testing/202505'

git_base_path <- '/u01/home/pbeati/2024/lucia/paper_acelerados/ncomms-24-47757'

```

### Load helpers

```{r helpers}

base_path <- file.path(git_base_path, 'accelerated_elements')
scripts_base_path <- file.path(base_path, 'source')
script_file_path <- file.path(scripts_base_path, 'common.R')

source(script_file_path)

```


## Data Preparation

### Alignments

The first step in the analysis is performed by the function `prepare_alignment_files` at <https://github.com/paulati/cladeAcc/blob/master/R/data_preparation.R>

This function downloads multiple alignment files from the public repository and checks their MD5 values.

In this example, we simply copy the sample MAF file to the working directory (either user-defined or the local temporary directory).
The function `copy_example_alignment` at xxxxxx performs this step.

```{r vignette_setup}

alignment_id_mammals <- '100_way'

copy_example_alignment(alignment_id_mammals, base_path)

```

### Neutral model

`prepare_neutral_model_file` downloads the neutral model from the public repository and verifies its MD5 checksum.

```{r neutral_model}

cladeAcc::prepare_neutral_model_file(alignment_id_mammals)

```
## Conservation

The `conserved_elements_computation` function at <https://github.com/paulati/cladeAcc/blob/master/R/conservation.R> returns the conserved elements for each clade, and the `conserved_elements_in_common` function returns their intersection.

The required species in each clade are evaluated by the functions `required_species_features_mammals_100way` and `required_species_features_sarcopterygii_100way`, found at <https://github.com/paulati/cladeAcc/blob/master/R/conservation_delegates.R>

Note: These functions can be customized with different criteria and passed via the `required_feats_func` parameter in `conserved_elements_computation`.


```{r conservation}

chr_mammals <- 114

clade_1 <- 'mammals'
chrs <- c(chr_mammals)
result_100_mammals <- cladeAcc::conserved_elements_computation(
    alignment_id_mammals, 
    clade_1, 
    chrs,
    required_feats_func = cladeAcc::required_species_features_mammals_100way)

print(result_100_mammals)

clade_2 <- 'sarcopterygii'
chrs <- c(chr_mammals)
result_100_sarc <- cladeAcc::conserved_elements_computation(
    alignment_id_mammals, 
    clade_2, 
    chrs,
    required_feats_func = cladeAcc::required_species_features_sarcopterygii_100way)

print(result_100_sarc)

cladeAcc::conserved_elements_in_common(alignment_id = alignment_id_mammals,
                                clade_1 = clade_1, 
                                clade_2 = clade_2, 
                                chr = chr_mammals,
                                conserved_path_1 = result_100_mammals[[1]], 
                                conserved_path_2 = result_100_sarc[[1]])



```

## Acceleration

To compute accelerated elements, the neutral model tree must include the 'MAMMALS' label.  This is added by the `label_neutral_model_tree_mammals` function. The `compute_accelerated_elements` then calculates acceleration scores for the candidate regions.

Note: The user can define their own `label_neutral_model_tree_mammals` function with different criteria and pass it via the `label_neutral_model_func` parameter to the `compute_accelerated_elements` function.


```{r acceleration}

acc_clade_mam <- 'mammals'
candidate_regions_clade_mam <- 'mammals_sarcopterygii'
chrs_mam <- list(chr_mammals)
branch_label_mam <- 'MAMMALS'
label_neutral_model_func_mam <- cladeAcc::label_neutral_model_tree_mammals

suppressWarnings(
    paths <- cladeAcc::compute_accelerated_elements(alignment_id_mammals, 
                                                acc_clade_mam,
                                                candidate_regions_clade_mam,
                                                chrs_mam, 
                                                branch_label_mam,
                                                label_neutral_model_func_mam,
                                                feat_lengths = c(25, 50)) 
)
                                             
print(paths)                                          

```
## Filtering

The scores computed in the previous step are filtered to obtain a set of conserved regions that overlap with accelerated elements.


```{r filtering}
    cladeAcc::process_filtering(alignment_id_mammals, acc_clade_mam, chrs_mam)
```

## Output

We define a "candidate functional element" as a conserved region (i.e., the set of candidate regions where acceleration was evaluated, obtained from  the previous step unified by a distance of less than 20 bp from each other and with a minimum size of 100 bp.

```{r output}
# read files:
score_base_path_base <- file.path(user_base_path, alignment_id_mammals, 'output',  
            'custom_filtering', 'mammals_sarcopterygii', 'candidate_elements',
            'score_filtered')
                             
score_base_path_base_25 <- file.path(score_base_path_base, '25')
score_base_path_base_50 <- file.path(score_base_path_base, '50')
file_name_25 <- paste0('chr', chr_mammals, '_score_25.csv')
file_name_50 <- paste0('chr', chr_mammals, '_score_50.csv')

file_path_25 <- file.path(score_base_path_base_25, file_name_25)
file_path_50 <- file.path(score_base_path_base_50, file_name_50)
            
file.exists(file_path_25)
file.exists(file_path_50)                 

data_25 <- read.delim(file_path_25, sep = '\t', header = TRUE)
data_50 <- read.delim(file_path_50, sep = '\t', header = TRUE)
data <- rbind(data_25, data_50)

data_sorted <- data[order(data$conserved_element_shift_count, decreasing = TRUE), ]

data_gr <- GenomicRanges::makeGRangesFromDataFrame(data_sorted, 
                                        seqnames.field = "phastCons_seqnames", 
                                        start.field = "phastCons_start", 
                                        end.field = "phastCons_end",  
                                        keep.extra.columns = TRUE)
                            
data_gr_reduced <- GenomicRanges::reduce(data_gr, min.gapwidth = 20L)

output <- data_gr_reduced[data_gr_reduced@ranges@width > 100]

output

```
---

# Acceleration in mammals (complete analysis)

To perform the complete analysis, the user must make a few minor changes:

1. Set the variables `chrs` and `chrs_mam` to vectors containing the chromosome ids to be processed. Example: `chrs <- c(14, 15, 16)`, `chrs_mam <- list(14, 15, 16)`  

2. Replace the sample alignment copy step with a call to `prepare_alignment_files` at <https://github.com/paulati/cladeAcc/blob/master/R/data_preparation.R>

3. Add a loop to compute `conserved_elements_in_common` for each chromosome.


---


